Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-20 paper-1/manuscript.tex
Rev sec Results:STD, F8 partial

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} Estimated STD-mediated depression $D$ has small direct effects.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the dev (left) and msc sequence (center), and their contrast (right). The color scale is shared across columns (and between dev and msc) for direct comparability.
\textbf{b} Statistics across networks and stimuli of the contrast (dev - msc) in $D$ in A trials (left) and B trials (right)

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The msc sequence, however, clearly evoked greater depression, as borne out in the contrast (dev - msc), which was negative throughout. Conversely, depression in B trials (bottom) were spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response size: The context was the same, but responses to B tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site (see also Figure \ref{fig6}a), with depression by definition similarly limited. As a result, the contrast (dev - msc) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicative of greatly increased synaptic depression in dev.

In order to show that this was a robust effect, we turn to statistics across networks and stimuli. We calculated the trial-average depression values $D^A$ and $D^B$ in dev and msc conditions, subtracted them, then calculated the mean over neurons, excluding neurons with less than 0.2 spikes per trial in both conditions [because $\Delta D^B$ $\approx$ zero if I don't do this. Not sure how to justify...]. The resulting contrasts are summarized in the left half of Figure \ref{fig8}b. We note first that, using the same exclusion criteria, we found that neurons received more depression in B trials than in A trials in the dev sequence (t = -4.4, p = 1.17e-05; one-sided t-test, n = 120), but not in the msc sequence (t = -1.14, p = 0.129). Furthermore, $D^B$ was greater in dev than msc (t = 2.94, p = 0.002). Both findings confirm that STD carried over between the frequent standard trials, but not, or less so, between control trials.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} Estimated STD-mediated depression $D$ has small direct effects.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the dev (left) and msc sequence (center), and their contrast (right). The color scale is shared across columns (and between dev and msc) for direct comparability.
\textbf{b} Statistics across networks and stimuli of the contrast (dev - msc) in $D$ in A trials (left) and B trials (right)
\textbf{c} Relationship across neurons of the sample network between the contrast (dev - msc) in $D^S$ and the contrast in average number of spikes fired in $S$ trials, and associated linear regression. Left, target trials; right, B trials.
\textbf{d} Network-wide mean contrast (dev - msc) in $D^S$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in panel c ($\Delta D^S$ vs response size contrast across neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided t-test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The msc sequence, however, clearly evoked greater depression, as borne out in the contrast (dev - msc), which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response size: The context was the same, but responses to B tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site (see also Figure \ref{fig6}a), with depression therefore similarly limited. As a result, the contrast (dev - msc) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicating greatly increased synaptic depression in dev.

To show how robust these effects were, we turn to statistics across networks and stimuli. We calculated the trial-average depression values $D^A$ and $D^B$ in dev and msc conditions, subtracted them, then calculated the mean over neurons. The resulting contrasts are summarized in Figure \ref{fig8}b. We note first that neurons were subject to more depression in B trials than in A trials in the dev sequence (-4.47, p = 9.02e-06; one-sided t-test, n = 120, data not shown), but not in the msc sequence (by definition). Furthermore, $D^B$ was greater in dev than msc (8.02, p = 4.05e-13). Both findings confirm that STD carried over between the frequent standard trials, but not, or less so, between the less frequent control and deviant trials.
However, we also found that $D^A$ was greater in msc than dev (-7.54, p = 5.04e-12; one-sided t-test, n = 120), suggesting that STD did exert stimulus-independent suppression in the msc sequence, as is clearly seen in the sample network, and similar to what we described earlier for TA.

Finally, to show that $D$ reflects suppression of a neuron's activity, we quantified the relationship between the contrasts in $D$ and the corresponding response size contrast. In the sample network (Figure \ref{fig8}c), we see that neurons with a greater decrease in $D$ from msc to dev had correspondingly greater response increase, and vice versa. Across both A and B trials, this relationship was captured well by a linear regression (A: Pearson's $\rho$ = -0.541, p = 2.28e-77, B: $\rho$ = -0.186, p = 1.61e-09; one-sided t-test, n = 1000). These findings were borne out across most networks and stimuli (Figure \ref{fig8}d), showing that $D$ captures the impact of STD not only on $V_m$, but also on the resulting output spikes.

To sum up, in the full model, STD acted in much the same way as TA, causing both a reduction of the response to the frequently presented standard stimuli (dev B) by adaptation, and an increase in the response to deviant stimuli (dev A) by relief from adaptation, relative to control. % ...
